---
title: "Data Analysis Final"
author: "Kai Hendershot"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
TaxiData <- read.table("C:/Users/kaise/Downloads/chicago_taxi.txt",header = TRUE)
library(psych)
library(magrittr)
library(car)
library(ggformula)
library(gmodels)
library(class)
library(caret)
```
# View the data

First I wanted to see the data to look at what type of variables I'm going to be
working with.
```{r}
View(TaxiData)
```

Next I explored and looked at the data to understand what I'll be working with and 
the association between variables as well as see if there was any type of pattern 
that stood out between the daytype, total, and payment type.
```{r}
pairs.panels(TaxiData[c("daytype", "seconds", "miles", "tips", "total")])

gf_point(daytype~total, data = TaxiData, shape = ~payment_type,
         size = 4, color = ~daytype, alpha = .5,
         position = position_dodge2(width = 0.5))%>%
  gf_refine(scale_color_brewer(palette = 'Set1'))
```
# Normalize the data

Then I wrote a function to normalize the data set. I had to save the columns I 
wanted to normalize into a variable because I kept getting an error when I tried
to list them all out and this was the solution. I then normalized the variables 
I needed and saved them into a data frame called norm_Taxi. I then viewed the data to 
make sure it all worked out.
```{r}
normie <- function(x){
  return((x-min(x))/(max(x)-min(x)))
}

make_these_norm <- c("seconds", "miles", "tips", "total", "speed")
TaxiData$daytype <- as.factor(TaxiData$daytype)

norm_taxi <- as.data.frame(lapply(TaxiData[,make_these_norm], normie))
View(norm_taxi)
```
# Split the data

I next split into training and test set
```{r}
taxi.train.ind <- sample(1:nrow(TaxiData), nrow(TaxiData)*.6)
taxi.training.set <- norm_taxi[taxi.train.ind,]
taxi.test.set <- norm_taxi[-taxi.train.ind,]
taxi.training.labels <- TaxiData[taxi.train.ind, "daytype"]
taxi.test.labels <- TaxiData[-taxi.train.ind, "daytype"]
```
# Run test and interpret

I used the k nearest neighbors function for my prediction where it got the 30
nearest neighbors and I ran this on my training and test set of data
```{r}
td_pred <- knn(train = taxi.training.set, test = taxi.test.set,
               cl = taxi.training.labels, k = 30)
```

I created a crosstable so I can evaluate the performance of my prediction
```{r}
CrossTable(x = taxi.test.labels, y = td_pred, prop.chisq = FALSE)
```
The model correctly predicted 'weekday' in approximately 99.2% of cases where the
actual label was 'weekday'. The model correctly p redicted 'weekend' in about 98.5%
of cases where the actual label was 'weekend'.The table shows the performance of
a classification model in predicting 'weekday' and 'weekend' categories.Higher 
percentages in the diagonal cells indicate accurate predictions by the model for
the respective categories.Lower percentages outside the diagonal cells suggest
misclassifications or incorrect predictions.


# Part 2

I first needed to make the daytype a factor so I could run the model I created
becasue with it being a text variable it wouldn't run
```{r}
TaxiData$daytype <- as.factor(TaxiData$daytype)
```

I split the data into 70/30 with a seed for reproducibility 
```{r}
set.seed(123)
train_indices <- createDataPartition(TaxiData$daytype, p = 0.7, list = FALSE)
taxi_training <- TaxiData[train_indices, ]
taxi_testing <- TaxiData[-train_indices, ]
```


I used the glm function with family = biinomial for logit regression. After that
I looked at the model to see what it came up with and the first few values of 
the fitted values.
```{r}
final.model <- glm(daytype~seconds+miles+tips+total+speed, data = taxi_training,
                 family = "binomial")

final.model
summary(final.model)
head(final.model$fitted.values)
```

There was some significance in speed, there was pretty high significance in tips,
and the intercept itself was seen as pretty significant as well.

I created a function for changing odds into probabilities because probabilities
are more useful than log odds then I applied it to the linear predictors 
```{r}
gen.prob.function <- function(x){
  A <- exp(x)
  prob <- A/(1+A)
  return(prob)
}
gen.prob <- sapply(final.model$linear.predictors, gen.prob.function)
```


Make linear predictors to log odds,and fitted values to probabilities
```{r}
head(final.model$fitted.values)
sum(abs(gen.prob - final.model$fitted.values) < .0000002)
gen.predict <- ifelse(gen.prob > .25, "weekend", "weekday")
```

Make a cross table to analyze results 
```{r}
CrossTable(x=gen.predict, y=taxi_training$daytype, prop.chisq = FALSE)
```
The model predicted 'weekday' in approximately 77.6% of cases among observations
where the actual label was 'weekday'. The model predicted 'weekend' in 
approximately 89.8% of cases among observations where the actual label was
'weekend'. Overall the model was pretty good at predicting. 




